### 2017-05-01 2017mergingmerges/section/01-Introduction.tex
typesetting to cut down

Suppose we have two input streams of numeric identifiers, and wish to perform some analysis on these identifiers. The identifiers from both streams arrive sorted, but may include duplicates. We wish to produce an output stream of unique identifiers from the first input stream, as well as produce the unique union of identifiers from both streams. Here is how we might write the source code, where @S@ is for @S@-tream.
\begin{code}
  uniquesUnion : S Nat -> S Nat -> (S Nat, S Nat)
  uniquesUnion sIn1 sIn2
   = let  sUnique = group sIn1
          sMerged = merge sIn1 sIn2
          sUnion  = group sMerged
     in   (sUnique, sUnion)
\end{code}

The @group@ operator detects groups of consecutive identical elements and emits a single representative, while @merge@ combines two sorted streams so that the output remains sorted. This example has a few interesting properties. Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ pulls values from @sIn1@ and @sIn2@ depends on the values themselves. If all values from @sIn1@ are smaller than the ones from @sIn2@, then @merge@ will pull all values from @sIn1@ before pulling any from @sIn2@, and vice versa. Secondly, although @sIn1@ occurs twice in the program, at runtime we only want to handle the elements of each stream once. To achieve this, the compiled program must coordinate between the two uses of @sIn1@, so that a new value is read only when both the @group@ and @merge@ operators are ready to receive it. Finally, as the stream length is unbounded, we cannot buffer an arbitrary number of elements read from either stream, or risk running out of local storage space.

For an implementation, we might try coding each of the operators as a separate concurrent process, and send each stream element using an intra-process communication mechanism. Developing such an implementation could be easy or hard, depending on what language features are available for concurrency. However, worrying about the \emph{performance tuning} of such a system, such as whether we need back-pressure to prevent buffers from being overrun, or how to chunk the stream data to reduce the amount of communication overhead, is invariably a headache.

We might instead define some sort of uniform sequential interface for data sources, with a single `pull' function that provides the next value in each stream. Each operator could be given this interface, so that the next value from each result stream is computed on demand. This approach is commonly taken with implementations of physical operators in database systems \cite{Graefe:Volcano}. However, the `pull only' model does not support operators with multiple outputs, such as our derived @uniquesUnion@ operator, at least not without unbounded buffering. Suppose a consumer pulls many elements from the @sUnique@ output stream. In order to perform the @group@ operation, the implementation needs to pull the corresponding source elements from the @sIn1@ input stream \emph{as well} as buffering an arbitrary number of them. It needs to buffer these elements because they are also needed to perform the @merge@ operation. When a consumer finally pulls from @sUnion@ we will be able to drain the buffer of @sIn1@ elements, but not before.

Instead, for a single threaded program, we want to perform \emph{stream fusion}, which takes the implied dataflow network and produces a simple sequential loop that gets the job done without requiring extra process-control abstractions and without requiring unbounded buffering. Sadly, existing stream fusion transformations cannot handle our example. As observed by \citet{kay2009you}, both pull-based and push-based fusion have fundamental limitations. Pull-based systems such as short-cut stream fusion~\cite{coutts2007stream} cannot handle cases where a particular stream or intermediate result is used by multiple consumers. We refer to this situation as a \mbox{\emph{split} --- in the} dataflow network of our example the flow from input stream @sIn1@ is split into both the @group@ and @merge@ consumers.

Push-based systems such as foldr/build fusion~\cite{gill1993short} cannot fuse our example either, because they do not support operators with multiple inputs. We refer to such a situation as a \emph{join} --- in our example the @merge@ operator expresses a join in the data-flow network. Some systems support both pull and push: data flow inspired array fusion using series expressions~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators. More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse our example, but requires the program to be rewritten to use explicitly polarized stream types.

Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject value-dependent operators with value dependent control flow such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}. The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but operates at a much lower level. The polyhedral model is based around affine loops, which makes it difficult to support filter-like operators such as @group@ and @merge@.

In this paper we present Machine Fusion, a new approach. Each operator is expressed as a restricted, sequential imperative program with commands that include both @pull@ for reading from an input stream and @push@ for writing to an output stream. We view each operator as a process in a concurrent process network, and the control flow of each process as a simple state machine. Our fusion transform then \emph{sequentializes} the concurrent process network into a single process, by choosing a particular interleaving of the operator code that requires no unbounded intermediate buffers. When the fusion transform succeeds we know it has worked. There is no need to inspect intermediate representations of the compiler to debug poor performance, which is a common problem in systems based on general purpose program transformations \cite{lippmeier2012:guiding}.

In summary, we make the following contributions:
\begin{itemize}
\item a process calculus for encoding infinite streaming programs (\S\ref{s:Processes});
\item an algorithm for fusing these processes, the first to support splits and joins (\S\ref{s:Fusion});
\item numerical results that demonstrate that the algorithm is well behaved when the number of fused processes is large. The size of the fused result program is not excessive. (\S\ref{s:Evaluation});
\item a formalization and proof of soundness for the core fusion algorithm in Coq (\S\ref{s:Proofs}).
\end{itemize}

>>>>

Suppose we have two input streams of numeric identifiers, and wish to perform some analysis on these identifiers. The identifiers from both streams are sorted, but may include duplicates. We wish to produce an output stream of unique identifiers from the first input stream, as well as produce the unique union of both streams. Here is how we might write the source code, where @S@ is for @S@-tream.
\begin{code}
  uniquesUnion : S Nat -> S Nat -> (S Nat, S Nat)
  uniquesUnion sIn1 sIn2
   = let  sUnique = group sIn1
          sMerged = merge sIn1 sIn2
          sUnion  = group sMerged
     in   (sUnique, sUnion)
\end{code}

The @group@ operator detects groups of consecutive identical elements and emits a single representative, while @merge@ combines two sorted streams so that the output remains sorted. This example has a few interesting properties. Firstly, the data-access pattern of @merge@ is \emph{value-dependent}, meaning that the order in which @merge@ pulls values from @sIn1@ and @sIn2@ depends on the values themselves: at each step, @merge@ must compare the values from both streams, and choose the stream with the smaller value to pull from.
Secondly, although @sIn1@ occurs twice in the program, at runtime we only want to handle the elements of each stream once. To achieve this, the compiled program must coordinate between the two uses of @sIn1@, so that a new value is read only when both the @group@ and @merge@ operators are ready to receive it. Finally, as the stream length is unbounded, we cannot buffer an arbitrary number of elements read from either stream, or risk running out of local storage space.

To implement this, we might write each operator as its own concurrent process, sending stream elements over intra-process channels.
Developing this could be easy or hard, depending on the available language features for concurrency.
However, the \emph{performance tuning} of such a system, such as whether we need back-pressure to prevent buffers from being overrun, or how to chunk stream data to reduce communication overhead, is invariably a headache.

We might instead define a uniform sequential interface for data sources, with a single `pull' function that provides the next value in each stream. Each operator could be given this interface, so that values from each stream are computed on demand. This approach is commonly taken in database systems \cite{Graefe:Volcano}. However, this `pull' model does not support operators with multiple outputs, such as our example, at least not without unbounded buffering. Suppose a consumer pulls many elements from the @sUnique@ output stream. In order to perform the @group@ operation, the implementation needs to pull the corresponding source elements from the @sIn1@ input stream \emph{as well} as buffering an arbitrary number of them. It needs to buffer these elements because they are also needed to perform the @merge@ operation. When a consumer finally pulls from @sUnion@ we will be able to drain the buffer of @sIn1@ elements, but not before.

Instead, for a single threaded program, we want to perform \emph{stream fusion}, which takes the implied dataflow network and produces a simple sequential loop that does not require extra process-control abstractions or unbounded buffering. Sadly, existing stream fusion transformations cannot handle our example. As observed by \citet{kay2009you}, both pull-based and push-based fusion have fundamental limitations. Pull-based systems such as short-cut stream fusion~\cite{coutts2007stream} cannot handle cases where a particular stream or intermediate result is used by multiple consumers. We refer to this situation as a \mbox{\emph{split} --- in the} dataflow network of our example the flow from input stream @sIn1@ is split into both the @group@ and @merge@ consumers.

Push-based systems such as foldr/build fusion~\cite{gill1993short} cannot fuse our example either, because they do not support operators with multiple inputs. We refer to this as a \emph{join} --- in our example the @merge@ operator expresses a join in the data-flow network. Some systems support both pull and push: data flow inspired array fusion using series expressions~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators. More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse our example, but requires the program to be rewritten to use explicitly polarized stream types.

Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject operators with value dependent control flow such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}.
The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but is based around affine loops, which makes it difficult to support value-dependent operators  such as @group@ and @merge@.

In this paper we present Machine Fusion, a new approach. Each operator is expressed as a restricted, sequential imperative program which \emph{pulls} from input streams, and \emph{pushes} to output streams. We view each operator as a process in a concurrent process network. Our fusion transform then \emph{sequentializes} the concurrent process network into a single process, by choosing a particular interleaving of the operator code that requires no unbounded intermediate buffers. When the fusion transform succeeds we know it has worked. There is no need to inspect intermediate representations of the compiler to debug poor performance, which is a common problem in systems based on general purpose program transformations \cite{lippmeier2012:guiding}.

In summary, we make the following contributions:
\begin{itemize}
\item a process calculus for infinite streaming programs (\S\ref{s:Processes});
\item a fusion algorithm, the first to support splits and joins (\S\ref{s:Fusion});
\item benchmarks showing significant performance gains (\S\ref{s:Benchmarks});
\item proof of soundness for the fusion algorithm in Coq (\S\ref{s:Proofs}).
\end{itemize}

### 2017-05-01 2017mergingmerges/section/02-Processes.tex
attempt to fit in sigconf 12p. format
typesetting to cut down

When the predicate is true, the instruction @(push sOut1 v A3 [ l = v, f = F ])@ pushes the value @v@ to the output stream @sOut1@ and proceeds to the instruction at label @A3@, once the variable @l@ is set to @v@ and @f@ to @F@ (False).

@@

\begin{figure}

\begin{center}
\begin{alltt}
           group 
             = \(\lambda\) (sIn1: Stream Nat) (sOut1: Stream Nat). 
               \(\nu\) (f: Bool) (l: Nat) (v: Nat) (A0..A3: Label).
\end{alltt}
\begin{code}
               process
                { ins:    { sIn1  }
                , outs:   { sOut1 }
                , heap:   { f = T, l = 0, v = 0 }
                , label:  A0
                , instrs: { A0 = pull sIn1 v          A1 []
                          , A1 = case (f || (l /= v)) A2 []  A3 []
                          , A2 = push sOut1 v         A3 [ l = v, f = F ]
                          , A3 = drop sIn1            A0 [] } }
\end{code}
\end{center}
\vspace{1em}
\includegraphics[scale=1.1]{figures/state-group.pdf}
\caption{The group combinator}
\label{fig:Process:Group}
\end{figure}

% -----------------------------------------------------------------------------
\subsection{Merging}
\begin{figure}
\begin{alltt}
               merge
                 = \(\lambda\) (sIn1: Stream Nat) (sIn2: Stream Nat) (sOut2: Stream Nat). 
                   \(\nu\) (x1: Nat) (x2: Nat) (B0..E2: Label).
\end{alltt}
\begin{code}
                   process
                    { ins:    { sM1, sM2 }
                    , outs:   { sM3 }
                    , heap:   { x1 = 0, x2 = 0 }
                    , label:  B0
                    , instrs: { B0 = pull sIn1  x1   B1 []
                              , B1 = pull sIn2  x2   C0 []
                              , C0 = case (x1 < x2)  D0 []  E0 []
                              , D0 = push sOut2 x1   D1 []
                              , D1 = drop sIn1       D2 []
                              , D2 = pull sIn1  x1   C0 []
                              , E0 = push sOut2 x2   E1 []
                              , E1 = drop sIn2       E2 []
                              , E2 = pull sIn2 x2    C0 [] } }
\end{code}

\medskip
\includegraphics[scale=1.1]{figures/state-merge.pdf}
\caption{The merge combinator}
\label{fig:Process:Merge}
\end{figure}

@@

We also associate each of the joint result states with information describing whether or not each source state machine has already pulled a value from each of its input streams. For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})). The result state @F0@ represents a combination of the two source states @A0@ and @B0@. As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both `sIn1' and `sIn2' map to `none'.

@@

\begin{figure}
\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}
\caption{Fusing pull instructions}
\label{fig:Fusion:Pulls}
\end{figure}

% -----------------------------------------------------------------------------
\subsubsection{Fusing Cases}
\begin{figure}
\includegraphics[scale=1.1]{figures/fuse-case-pull.pdf}
\caption{Fusing case instructions}
\label{fig:Fusion:Case}
\end{figure}

@@

Fig.~\ref{fig:Process:Fused} shows the final result of fusing @group@ and @merge@ together. There are similar rules for handling the other combinations of instructions, but we defer the details to \S\ref{s:Fusion}. The result process has two input streams, @sIn1@ and @sIn2@ and two output streams: @sOut1@ which is the @group@ processes output stream, and @sOut2@ which is the @merge@ processes output stream.

Overall, our fusion algorithm has taken two separate processes that we once imagined to be running concurrently, and has produced a single sequential result process that implements both. We have chosen a \emph{specific sequential order} in which to interleave the instructions that implement both source processes. As with the Flow Fusion system of \citet{lippmeier2013data} we have performed the job of a concurrent scheduler at compile time. However, in contrast to Flow Fusion and similar systems, we do not need to organize statements into a fixed \emph{loop anatomy}, we simply merge them as they are. This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams, which we discuss further in \S\ref{s:FutureWork}.

To complete the implementation of our example from \S\ref{s:Introduction} we would now proceed to fuse the process from the final line (also a @group@) into this new result process. The order in which pairs of processes are fused together does matter, as does the order in which the instructions are interleaved --- we discuss both points further in \S\ref{s:Evaluation}.

@@

%%% AR: would like to highlight which machine is performing the current instruction, eg bolding "A0" when it's group moving from A0 to A1
\begin{figure}
\definecolor{groupc}{HTML}{308030}
\definecolor{mergec}{HTML}{800000}
\definecolor{sharec}{HTML}{000080}

@@

\caption{Fusion of \textcolor{groupc}{group} and \textcolor{mergec}{merge}, along with \textcolor{sharec}{shared} instructions}
\label{fig:Process:Fused}
\end{figure}

>>>>

% AR: Clumsy, but added "at L2 executes:" to force @(push...)@ onto its own line. Otherwise it goes off the right.
When the predicate is true, the instruction at label @A2@ executes:
@(push sOut1 v A3 [ l = v, f = F ])@
pushes the value @v@ to the output stream @sOut1@ and proceeds to the instruction at label @A3@, once the variable @l@ is set to @v@ and @f@ to @F@ (False).

@@

\begin{figure*}

\begin{minipage}{0.6\textwidth}
\begin{alltt}
 group 
   = \(\lambda\) (sIn1: Stream Nat) (sOut1: Stream Nat). 
     \(\nu\) (f: Bool) (l: Nat) (v: Nat) (A0..A3: Label).
\end{alltt}
\begin{code}
     process
      { ins:    { sIn1  }
      , outs:   { sOut1 }
      , heap:   { f = T, l = 0, v = 0 }
      , label:  A0
      , instrs: { A0 = pull sIn1 v          A1 []
                , A1 = case (f || (l /= v)) A2 []  A3 []
                , A2 = push sOut1 v         A3 [ l = v, f = F ]
                , A3 = drop sIn1            A0 [] } }
\end{code}
\end{minipage}
\begin{minipage}{0.29\textwidth}
\includegraphics[scale=1.0]{figures/state-group.pdf}
\end{minipage}

\caption{The group combinator}
\label{fig:Process:Group}
\end{figure*}

% -----------------------------------------------------------------------------
\subsection{Merging}
\begin{figure*}
\begin{alltt}
 merge
   = \(\lambda\) (sIn1: Stream Nat) (sIn2: Stream Nat) (sOut2: Stream Nat). 
     \(\nu\) (x1: Nat) (x2: Nat) (B0..E2: Label).
\end{alltt}
\begin{code}
     process
      { ins:    { sM1, sM2 }
      , outs:   { sM3 }
      , heap:   { x1 = 0, x2 = 0 }
      , label:  B0
      , instrs: { B0 = pull sIn1  x1   B1 []             , B1 = pull sIn2  x2   C0 []
                , C0 = case (x1 < x2)  D0 []  E0 []      , D0 = push sOut2 x1   D1 []
                , D1 = drop sIn1       D2 []             , D2 = pull sIn1  x1   C0 []
                , E0 = push sOut2 x2   E1 []             , E1 = drop sIn2       E2 []
                , E2 = pull sIn2 x2    C0 [] } }
\end{code}
% \begin{minipage}{0.29\textwidth}
\includegraphics[scale=1.0]{figures/state-merge.pdf}
% \includegraphics[scale=1.0,angle=-90,origin=c]{figures/state-merge.pdf}
% \end{minipage}

\caption{The merge combinator}
\label{fig:Process:Merge}
\end{figure*}

@@

We also associate each of the joint result states with whether or not each source machine has already pulled a value from each of its input streams. For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{sIn1 = none\}), (B0, \{sIn1 = none, sIn2 = none\})). The result state @F0@ represents a combination of the two source states @A0@ and @B0@. As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both `sIn1' and `sIn2' map to `none'.

@@

\begin{figure*}
\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}
\caption{Fusing pull instructions}
\label{fig:Fusion:Pulls}
\end{figure*}

% -----------------------------------------------------------------------------
\subsubsection{Fusing Cases}
\begin{figure*}
\includegraphics[scale=1.1]{figures/fuse-case-pull.pdf}
\caption{Fusing case instructions}
\label{fig:Fusion:Case}
\end{figure*}

@@

Fig.~\ref{fig:Process:Fused} shows the final result of fusing @group@ and @merge@ together. There are similar rules for handling the other combinations of instructions, but we defer the details to \S\ref{s:Fusion}. The result process has two input streams, @sIn1@ and @sIn2@ and two output streams: @sOut1@ which is @group@'s output stream, and @sOut2@ which is @merge@'s output stream.

Overall, our fusion algorithm has taken two separate processes that we once imagined to be running concurrently, and has produced a single sequential result process that implements both. We have chosen a \emph{specific sequential order} in which to interleave the instructions that implement both source processes. As with Flow Fusion \cite{lippmeier2013data} we have performed the job of a concurrent scheduler at compile time. However, in contrast to Flow Fusion and similar systems, we do not need to organize statements into a fixed \emph{loop anatomy}, we simply merge them as they are. This allows us to implement a wider range of processes, including ones with nested loops that work on segmented streams, which we discuss further in \S\ref{s:FutureWork}.

To complete the implementation of our example from \S\ref{s:Introduction} we would now proceed to fuse the process from the final line (also a @group@) into this new result process. The order in which pairs of processes are fused together does matter, as does the order in which instructions are interleaved --- we discuss both points further in \S\ref{s:Evaluation}.

@@

%%% AR: would like to highlight which machine is performing the current instruction, eg bolding "A0" when it's group moving from A0 to A1
\begin{figure*}
\definecolor{groupc}{HTML}{308030}
\definecolor{mergec}{HTML}{800000}
\definecolor{sharec}{HTML}{000080}

@@

\caption{Fusion of \textcolor{groupc}{group} and \textcolor{mergec}{merge}, along with \textcolor{sharec}{shared} instructions}
\label{fig:Process:Fused}
\end{figure*}

### 2017-05-01 2017mergingmerges/section/03-Definition.tex
typesetting to cut down

\begin{enumerate}
\item \emph{Injection} of a single value from a stream into a process, or a set of processes. Each individual process only needs to accept the value when it is ready for it, and injection of a value into a set of processes succeeds only when they \emph{all} accept it.

@@

Execution of a process network is non-deterministic. At any moment several processes may be able to take a step, while others are blocked. As with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking. This enables the overall sequence of values on each output channel to be deterministic. Unlike Kahn processes, pushing to a channel can also block. Each consumer has a single element buffer, which is the stream buffer variable, and pushing can only succeed when that buffer is empty.

@@

Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions, is not. When we fuse two processes we exploit this fact by choosing one particular instruction ordering that enables the process network to advance without requiring unbounded buffering.

Each output channel may be pushed to by a single process only, so in a sense each output channel is owned by a single process. The only intra-process communication is via channels and streams. Our model is ``pure data flow'' (or perhaps ``functional data flow'') as there are no side-channels between the processes. This is in contrast to systems such as StreamIt~\cite{thies2002streamit}, where the processes are also able to send asynchronous messages to each other, in addition to via the formal input and output streams.

% -----------------------------------------------------------------------------
\subsubsection{Injection}
Fig.~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes. The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields an updated process $p'$''. The @injects@ form is similar, but operates on sets of processes rather than a single one.

Rule (InjectValue) injects of a single value into a single process. The value is stored as a (@pending@~ v) binding in the $\InputState$ of the associated channel of the process. The $\InputState$ acts as a single element buffer, and must be empty (set to @none@) for the injection to succeed.

>>>>

\begin{enumerate}
\item \emph{Injection} of a single value from a stream into a process, or a set of processes. Each individual process only needs to accept the value when it is ready for it, and injection into a set of processes succeeds only when they \emph{all} accept it.

@@

Execution of a process network is non-deterministic. At any moment several processes may be able to take a step, while others are blocked. As with Kahn processes~\cite{kahn1976coroutines}, pulling from a channel is blocking. This enables the overall sequence of values on each output channel to be deterministic. Unlike Kahn processes, pushing to a channel can also block. Each consumer has a single element buffer, and pushing only succeeds when that buffer is empty.

@@

Importantly, it is the order in which values are \emph{pushed to each particular output channel} which is deterministic, whereas the order in which different processes execute their instructions is not. When we fuse two processes we exploit this fact by choosing one particular instruction ordering that enables the process network to advance without requiring unbounded buffering.

Each channel may be pushed to by a single process only, so in a sense each output channel is owned by a single process. The only intra-process communication is via channels and streams. Our model is ``pure data flow'' as there are no side-channels between processes. This is in contrast to systems such as StreamIt~\cite{thies2002streamit}, where processes can communicate via asynchronous messages as well as channels.

% -----------------------------------------------------------------------------
\subsubsection{Injection}
Fig.~\ref{fig:Process:Eval:Inject} gives the rules for injecting values into processes. The statement $(\ProcInject{p}{v}{c}{p'})$ reads ``given process $p$, injecting value $v$ into channel $c$ yields an updated process $p'$''. The @injects@ form is similar, operating on sets of processes.

Rule (InjectValue) injects a single value into a single process. The value is stored as a (@pending@~ v) binding in the $\InputState$ of the associated channel of the process. The $\InputState$ acts as a single element buffer, and must be empty (@none@) for injection to succeed.
